---
title: "Cars and Crime Symposium Workshop"
subtitle: "Accessing and Analyzing UK Road Crash Data with stats19"
date: last-modified
---

Welcome to the **Cars and Crime Symposium Workshop** repository!

- **Online Slides:** [robinlovelace.github.io/carscrime/slides.html](https://robinlovelace.github.io/carscrime/slides.html)
- **Source Code:** [github.com/Robinlovelace/carscrime](https://github.com/Robinlovelace/carscrime)

![Deploy status](https://img.shields.io/github/actions/workflow/status/Robinlovelace/carscrime/publish.yml?branch=main&label=deploy) Last updated: {{< meta date >}}

## Casualties over time

```{r}
#| message: false
#| warning: false
library(stats19)
library(tidyverse)
library(readr)
```

```{r}
#| message: false
# Load pre-aggregated STATS19 dataset
cas_by_year_type <- read_csv("data/casualties_by_year_type.csv", show_col_types = FALSE)
```

```{r}
# Absolute casualties over time
ggplot(cas_by_year_type, aes(x = collision_year, y = n, color = type)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 1.8) +
  scale_color_brewer(palette = "Set1") +
  labs(
    x = "Year", y = "Number of casualties", color = "Casualty Type",
    title = "UK Road Casualties by Transport Mode (1979 - Present)"
  ) +
  theme_minimal()
```

```{r}
# Indexed plot (1979 = 100)
cas_1979 <- cas_by_year_type |>
  filter(collision_year == 1979) |>
  select(type, n) |>
  rename(n_1979 = n)

cas_by_year_type |>
  left_join(cas_1979, by = "type") |>
  mutate(n_indexed = n / n_1979 * 100) |>
  ggplot(aes(x = collision_year, y = n_indexed, color = type)) +
  geom_line(linewidth = 1.2) +
  scale_color_brewer(palette = "Set1") +
  labs(
    x = "Year", y = "Casualties indexed (1979 = 100)", color = "Casualty Type",
    title = "Relative Change in Casualties Since 1979"
  ) +
  theme_minimal()
```

## Car harms: How has the burden shifted?

The plots below illustrate how "car harms" have evolved differently for people **driving**, **walking**, and **cycling**.

### Car occupant casualties — safety technology at work

Car occupant casualties have fallen dramatically since 1979, driven by seatbelt laws, airbags, crumple zones, and improved vehicle crash protection.

```{r}
cas_by_year_type |>
  filter(type == "Car occupant") |>
  ggplot(aes(x = collision_year, y = n)) +
  geom_line(color = "#1F78B4", linewidth = 1.2) +
  geom_point(color = "#1F78B4", size = 2) +
  labs(
    x = "Year", y = "Car occupant casualties",
    subtitle = "Dramatic decline as vehicle safety improves for occupants inside cars"
  ) +
  theme_minimal()
```

### Pedestrians and Cyclists hit by cars: Hurt and Killed since 1979

By joining casualty data with vehicle table records, we isolate vulnerable road users (pedestrians and cyclists) struck in collisions involving cars, tracking how many were **killed** versus **hurt (injured)** each year since 1979.

```{r}
#| message: false
vru_car_harm <- read_csv("data/vru_car_harm.csv", show_col_types = FALSE)

ggplot(vru_car_harm, aes(x = collision_year, y = n, color = outcome)) +
  geom_line(linewidth = 1.2) +
  facet_wrap(~ vru_mode, scales = "free_y") +
  scale_color_manual(values = c("Killed" = "#D95F02", "Hurt (Injured)" = "#7570B3")) +
  labs(
    x = "Year", y = "Number of casualties",
    title = "Pedestrians and Cyclists Hit by Cars (1979 - Present)",
    subtitle = "Annual counts of vulnerable road users killed vs hurt in car collisions",
    color = "Outcome"
  ) +
  theme_minimal()
```

### Risk shift: vulnerable road user casualties per car occupant

This ratio shows for every car occupant casualty, how many pedestrians and cyclists are hurt. A rising ratio indicates that the **burden of risk has shifted** onto people *outside* cars.

```{r}
ratio_data <- cas_by_year_type |>
  filter(type %in% c("Car occupant", "Pedestrian", "Cyclist")) |>
  pivot_wider(names_from = type, values_from = n) |>
  mutate(
    ped_per_car = Pedestrian / `Car occupant`,
    cycle_per_car = Cyclist / `Car occupant`
  )

ggplot(ratio_data, aes(x = collision_year)) +
  geom_line(aes(y = ped_per_car, color = "Pedestrians per car occupant"), linewidth = 1.2) +
  geom_point(aes(y = ped_per_car, color = "Pedestrians per car occupant"), size = 1.8) +
  geom_line(aes(y = cycle_per_car, color = "Cyclists per car occupant"), linewidth = 1.2) +
  geom_point(aes(y = cycle_per_car, color = "Cyclists per car occupant"), size = 1.8) +
  scale_color_manual(values = c("Pedestrians per car occupant" = "#E7298A",
                                 "Cyclists per car occupant" = "#66A61E")) +
  labs(
    x = "Year", y = "Casualty ratio",
    color = NULL,
    subtitle = "Risk burden has shifted from car occupants to vulnerable road users"
  ) +
  theme_minimal()
```

## Additional Trends in Car Harms

### Trend 1: Car size & engine capacity inflation (`engine_capacity_cc`) over time

As cars have grown heavier, wider, and more powerful (the "SUV-ification" of passenger vehicles), average engine displacement of cars involved in collisions with pedestrians and cyclists has steadily increased.

```{r}
#| message: false
car_engine_vru <- read_csv("data/car_engine_vru.csv", show_col_types = FALSE)

ggplot(car_engine_vru, aes(x = collision_year)) +
  geom_line(aes(y = mean_cc, color = "Mean engine capacity (cc)"), linewidth = 1.2) +
  geom_line(aes(y = median_cc, color = "Median engine capacity (cc)"), linewidth = 1.2, linetype = "dashed") +
  scale_color_manual(values = c("Mean engine capacity (cc)" = "#D95F02",
                                 "Median engine capacity (cc)" = "#1B9E77")) +
  labs(
    x = "Year", y = "Engine capacity (cc)",
    color = "Metric",
    title = "Engine Capacity of Cars Involved in Vulnerable Road User Collisions",
    subtitle = "Steady rise in engine size of cars striking pedestrians and cyclists"
  ) +
  theme_minimal()
```

### Trend 2: Driver Sex in Vulnerable Road User Collisions

Analyzing vehicle driver attributes shows a persistent, large disparity in driver sex for cars involved in collisions striking pedestrians and cyclists.

```{r}
#| message: false
driver_sex_vru <- read_csv("data/driver_sex_vru.csv", show_col_types = FALSE)

ggplot(driver_sex_vru, aes(x = collision_year, y = n, color = sex_of_driver)) +
  geom_line(linewidth = 1.2) +
  scale_color_manual(values = c("Male" = "#377EB8", "Female" = "#E41A1C")) +
  labs(
    x = "Year", y = "Number of drivers in VRU collisions",
    color = "Driver Sex",
    title = "Driver Sex in Car Collisions Involving Pedestrians and Cyclists",
    subtitle = "Male drivers account for the majority of cars involved in VRU collisions"
  ) +
  theme_minimal()
```

{{< include _readme-body.md >}}
