---
title: "Cars and Crime Symposium Workshop"
subtitle: "Accessing and Analyzing UK Road Crash Data with stats19"
date: last-modified
---

Welcome to the **Cars and Crime Symposium Workshop** repository!

- **Online Slides:** [robinlovelace.github.io/carscrime/slides.html](https://robinlovelace.github.io/carscrime/slides.html)
- **Source Code:** [github.com/Robinlovelace/carscrime](https://github.com/Robinlovelace/carscrime)

![Deploy status](https://img.shields.io/github/actions/workflow/status/Robinlovelace/carscrime/publish.yml?branch=main&label=deploy) Last updated: {{< meta date >}}

## Casualties over time

```{r}
#| cache: true
library(stats19)
library(tidyverse)
library(readr)
```

```{r}
#| cache: true
# Read pre-aggregated casualty data (built from raw STATS19 by build-data.R)
casualties_by_year_type <- read_csv("data/casualties_by_year_type.csv",
                                    show_col_types = FALSE)
```

```{r}
# Absolute casualties over time
ggplot(casualties_by_year_type, aes(x = collision_year, y = n, color = type)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2) +
  labs(x = "Year", y = "Number of casualties", color = "Type") +
  theme_minimal()
```

```{r}
# Indexed plot (1979 = 100)
collisions_1979 <- casualties_by_year_type |>
  filter(collision_year == 1979) |>
  select(type, n) |>
  rename(n_1979 = n)

casualties_by_year_type |>
  left_join(collisions_1979, by = "type") |>
  mutate(n_indexed = n / n_1979 * 100) |>
  ggplot(aes(x = collision_year, y = n_indexed, color = type)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2) +
  labs(x = "Year", y = "Casualties indexed (1979 = 100)", color = "Type") +
  theme_minimal()
```

## Car harms: How has the burden shifted?

The three plots below illustrate how "car harms" have evolved differently for people **driving**, **walking**, and **cycling**.

### Car occupant casualties — safety technology at work

Car occupant casualties have fallen dramatically since 1979, driven by seatbelt laws, airbags, crumple zones, and improved vehicle design.

```{r}
casualties_by_year_type |>
  filter(type == "Car occupant") |>
  ggplot(aes(x = collision_year, y = n)) +
  geom_line(color = "#1F78B4", linewidth = 1.2) +
  geom_point(color = "#1F78B4", size = 2) +
  labs(x = "Year", y = "Car occupant casualties",
       subtitle = "Dramatic decline as vehicle safety improves") +
  theme_minimal()
```

### Pedestrian casualties — harm to people walking

While car occupants have become dramatically safer, pedestrian casualties have also declined — but not nearly as steeply, and from a much higher baseline per trip.

```{r}
casualties_by_year_type |>
  filter(type == "Pedestrian") |>
  ggplot(aes(x = collision_year, y = n)) +
  geom_line(color = "#E7298A", linewidth = 1.2) +
  geom_point(color = "#E7298A", size = 2) +
  labs(x = "Year", y = "Pedestrian casualties",
       subtitle = "Slower progress for people outside cars") +
  theme_minimal()
```

### Risk shift: vulnerable road user casualties per car occupant

This ratio tells the story: for every car occupant casualty, how many pedestrians and cyclists are hurt? A rising ratio means the **burden of risk has shifted** from people *inside* cars to people *outside* them.

```{r}
ratio_data <- casualties_by_year_type |>
  filter(type %in% c("Car occupant", "Pedestrian", "Cyclist")) |>
  pivot_wider(names_from = type, values_from = n) |>
  mutate(
    ped_per_car = Pedestrian / `Car occupant`,
    cycle_per_car = Cyclist / `Car occupant`
  )

ggplot(ratio_data, aes(x = collision_year)) +
  geom_line(aes(y = ped_per_car, color = "Pedestrians per car occupant"), linewidth = 1.2) +
  geom_point(aes(y = ped_per_car, color = "Pedestrians per car occupant"), size = 2) +
  geom_line(aes(y = cycle_per_car, color = "Cyclists per car occupant"), linewidth = 1.2) +
  geom_point(aes(y = cycle_per_car, color = "Cyclists per car occupant"), size = 2) +
  scale_color_manual(values = c("Pedestrians per car occupant" = "#E7298A",
                                 "Cyclists per car occupant" = "#66A61E")) +
  labs(x = "Year", y = "Casualty ratio",
       color = NULL,
       subtitle = "Risk burden has shifted from car occupants to vulnerable road users") +
  theme_minimal()
```

{{< include _readme-body.md >}}
